---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(tidyverse)
library(alohaal)
```

# alohaal

<!-- badges: start -->
<!-- badges: end -->

The goal of alohaal is to ...

## Installation


And the development version from [GitHub](https://github.com/) with:

```{r}
# install.packages("devtools")
#devtools::install_github("adhil0/alohaal")
```
## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
#library(alohaal)
## basic example code
```

| Dataset     | Description                                                                             | Techniques Used  |
| ----------- | -----------                                                                             | ---------------- |
| nwt_pikas   | Pika demography data for west knoll and Indian Peaks wilderness, 2008 - ongoing         |ggmap, sf,  ggplot2, spatial plotting |
| ntl_ice   | North Temperate Lakes LTER: Ice Duration - Madison Lakes Area 1853 - current        | joining datasets, Time Series, line plots                 |
